---
format: live-html
engine: knitr
code-annotations: below
---

# R as a calculator

:::::: columns
::: {.column width="49%"}
**Maths:** `+ - * / ^`

```{r}
3 + 1 - 2

(3 + 4) / 2

3^2

# Division 
15 / 6
15 %/% 6 # integer part
15 %% 6 # remainder

```
:::

::: {.column width="2%"}
<!-- empty column to create gap -->
:::

::: {.column width="49%"}
**Assign:** `<- =`

```{r}
x <- 3 # <1>
y = 3 

```

1.  **Note:** In R, both `<-` and `=` can be used to assign a value to an object. `<-`is the conventional assignment operator in R. You will often see `=` used inside function calls to specify arguments. Keeping these two uses separate can make code easier to read.

**Comparison:** `== != < > <= >=`

```{r}
y == x 
x == 6
y != x 
```

**Logical:** `&` and `|`

```{r}
(x < 5) & (y == 5)
(x < 5) | (y == 5)
```
:::
::::::

## Objects

-   **vectors**

-   **data frames** (and tibbles)

-   matrices (and arrays)

-   lists

## Vectors

-   one-dimensional collection of values

Common vector types/classes:

-   logical: TRUE / FALSE

-   numeric: a vector that contains numbers

-   character: a vector that contains strings of text

```{r}
logical_vector   <- c(TRUE, FALSE, TRUE)
numeric_vector   <- c(4, 5.2, pi)
character_vector <- c("four", "five", "six")
```

More info see: [r-bloggers](https://www.r-bloggers.com/2023/09/understanding-data-types-in-r/)

```{r}
# c() creates a vector
v <- c(4, "four", 1+pi) # <1>

# [ ] lets you subset from the vector 
v[3]
```

1.  Atomic vectors can only contain one type. So adding a character to your vector makes all the values characters (coercion)

::: callout-note
```{r}
v_character <- c("four", 1+pi, 1.5+exp(1), 'almost four')# "" vs '' is preference
is.character(v_character)
class(v_character) 
typeof(v_character) 
```

```{r}
v_numeric <- c(4, 1+pi, 1.5+exp(1))
is.numeric(v_numeric)
class(v_numeric) # high-level data structure - user's perspective
typeof(v_numeric) # internal representation
```

returns "double" because by default numeric values are stored as "double-precision numbers"
:::

# Exercises

### Exercise 1: Debug the code

Each line below contains a mistake. Correct it.

```{webr}
x <- c(1 2 3)

cities <- c(Amsterdam, Rotterdam, Utrecht)

x =< 5

c(1, 2, 3)[0]
```

