# Practicalities

## File types

### Scripts

An R script is a plain text file that contains R code. R scripts use the file extension `.R`.

You can run:

-   a single line of code
-   several selected lines
-   the entire script

Comments start with `#`. R ignores anything after `#` on that line.

Comments are useful for explaining what your code does or leaving notes for yourself.

### Notebooks

A notebook (Rmarkdown `.Rmd` or Quarto `.qmd`) is a file that can contain both normal text and executable R code.

Unlike an R script, a notebook can combine:

-   explanatory text
-   R code
-   output
-   tables
-   figures
-   equations

R code is placed inside **code chunks**

When you run the chunk, the output can appear directly underneath the code.

This makes notebooks particularly useful for data analysis because you can keep the **code, results, and explanation together in one document**.

## Packages

R comes with many built-in functions, but you can add more functionality using **packages**.

### Installing packages

A package only needs to be installed once:

``` r
install.packages("tidyverse")
```

### Loading packages

A package needs to be loaded each time you start a new R session:

``` r
library(tidyverse)
```

## Functions

Functions perform actions in R. You have already used functions such as `c()`.

A function is usually followed by parentheses:

```{r}
mean(c(1, 2, 3))
```

Values supplied to a function are called **arguments**:

```{r}
round(3.14159, digits = 2)
```

Packages provide additional functions. For example, after loading `ggplot2` from `tidyverse`, you can use functions provided by that package.

## Writing readable R code

Some simple habits make your R code much easier to read:

-   use meaningful object names
-   add spaces around operators
-   use comments where they are useful
-   keep your formatting consistent

See: [R-bloggers](https://www.r-bloggers.com/2018/09/r-code-best-practices/)
