---
format: live-html
engine: knitr
live:
  show-hints: true
  show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

# Working with data

## Functions

R has a set of descriptive statistics functions.

| Function    | Description                        |
|-------------|------------------------------------|
| `table(x)`  | Counts how often each value occurs |
| `sum(x)`    | Adds the values                    |
| `mean(x)`   | Arithmetic mean                    |
| `median(x)` | Median                             |
| `var(x)`    | Variance                           |
| `sd(x)`     | Standard deviation                 |
| `min(x)`    | Minimum                            |
| `max(x)`    | Maximum                            |
| `range(x)`  | Returns the minimum and maximum    |

```{r}
# Example
marks <- c(7.5, 9, NA, 8, 6.5, 8, 7.5)
round(sd(marks, na.rm = TRUE), 3)
range(marks, na.rm = TRUE)
```

::: callout-important
## Missing values

Many functions return `NA` when the data contain missing values because the functions have `na.rm = TRUE` by default. Use `na.rm = TRUE` when you want the calculation to ignore missing values.

```{r}
mean(marks)
mean(marks, na.rm = TRUE)
```
:::

```{r}
table(marks)
```

::: callout-note
`table()` does not show missing values by default. If you want to see them, try:

```{r}
table(marks, useNA = "ifany")
```
:::

## Your turn: the `iris` data

The built-in `iris` data frame contains measurements for 150 iris flowers. You can inspect its help page with `?iris`.

### Exercise 1 — What is in the data frame?

How many variables does `iris` contain? What are their names?

```{webr}
#| setup: true
#| exercise: iris_1
data(iris)
```

```{webr}
#| exercise: iris_1
# Number of variables
______

# Variable names
______
```

:::: {.hint exercise="iris_1"}
::: {.callout-note collapse="false"}
## Hint 1

Two useful functions are `ncol()` and `names()`.
:::
::::

:::: {.solution exercise="iris_1"}
::: {.callout-tip collapse="false"}
## Fully worked solution

```{r}
ncol(iris)
names(iris)
```
:::
::::

### Exercise 2 — Count flowers by species

Create a frequency table showing how many flowers belong to each species.

```{webr}
#| exercise: iris_2
table(iris$______)
```

:::: {.hint exercise="iris_2"}
::: {.callout-note collapse="false"}
## Hint 1

Use the `Species` column.
:::
::::

:::: {.hint exercise="iris_2"}
::: {.callout-note collapse="false"}
## Hint 2

Use `$` to select a column from a data frame
:::
::::

:::: {.solution exercise="iris_2"}
::: {.callout-tip collapse="false"}
## Fully worked solution

```{r}
table(iris$Species, useNA = "ifany")
```
:::
::::

### Exercise 3 — Summarise petal length

Compute the **mean**, **median**, and **range** of `Petal.Length`.

```{webr}
#| exercise: iris_3

```

:::: {.hint exercise="iris_3"}
::: {.callout-note collapse="false"}
## Hint 1

You need the functions `mean()`, `median()`, and `range()`.
:::
::::

:::: {.hint exercise="iris_3"}
::: {.callout-note collapse="false"}
## Hint 2

Select petal length with: `iris$Petal.Length`
:::
::::

:::: {.solution exercise="iris_3"}
::: {.callout-tip collapse="false"}
## Fully worked solution

```{r}
mean(iris$Petal.Length)
median(iris$Petal.Length)
range(iris$Petal.Length)
```
:::
::::

# Data wranging

Tidyverse vs Base R

```{r baseR_ifelse}
dat <- readRDS("data/PlannedEnv_4countries.rds")

                # ifelse(test, Yes, No)
dat$continent <- ifelse(dat$country == "US" | dat$country == "Mexico", 
                        "N. America",
                        ifelse(dat$country == "India", 
                               "Asia", 
                               ifelse(dat$country == "Spain",
                                      "Europe",
                                      NA)))

```

If country is US or Mexico, then its N. America, if it is India then Asia, if Spain then Europe, if it is none these then NA

```{r tidy_ifelse}
library(tidyverse)

dat <- dat |>
  mutate(continent_tidy = case_when(
    country %in% c("US", "Mexico") ~ "N. America",
    country == "India" ~ "Asia",
    country == "Spain" ~ "Europe",
    is.na(country) ~ NA
  ))

any(dat$continent == dat$continent_tidy)
```

```{r}
dat_female <- subset(dat, gender == "female")

dat_female <- dat|>
  filter(gender == "female")

```

# Exercises

For exercises [Applied Data Skills](https://psyteachr.github.io/ads-v3/), chapters:

-   4 Data Summaries

-   9 Data Wrangling
