---
format: live-html
engine: knitr
live:
  show-hints: true
  show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

# Data Visualization

```{r setup, echo = FALSE}

source("_setup.R")

```

### Why plot?

```{r echo = FALSE}
data("datasaurus_dozen")
datasaurus_dozen |>
  subset(dataset == "dino"| dataset == "star") |>
  group_by(dataset) |>
  summarise(mean_x = mean(x),
            mean_y = mean(y),
            sd_x = sd(x),
            sd_y = sd(y),
            n(),
            correlation = round(cor(x, y),2)) |>
  kable(col.names = c("Dataset", "Mean x", "Mean y",
                      "SD x", "SD y", "Sample size", "Correlation (x,y)")) |>
  kable_styling()
            

```

Based only on this table, we might conclude that the datasets tell the same story.

::: {.callout-caution collapse="true"}
```{r, echo=FALSE}
datasaurus_dozen |>
  subset(dataset == "dino"| dataset == "star") |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  facet_wrap(~dataset) +
  theme_bw()
```

A numerical summary answers a specific question. A plot helps us notice things we did not think to ask.
:::

## Start with your question

| Question | Data pattern | Base R | GGplot2 |
|------------------|------------------|------------------|-------------------|
| How much? | Categories and amounts | `barplot()` | `geom_col()` or `geom_bar()` |
| How is it distributed? | One quantitative variable | `hist()` or `density()` | `geom_histogram()` or `geom_density()` |
| Are variables related? | Two quantitative variables | `plot()` | `geom_point()` |
| How does it change? | Ordered or time-based observations | `plot(type = "l")` | `geom_line()` |
| Does the pattern differ? | A variable defining groups | `subset()` or `split()` | `facet_wrap()` |

# Base R

### Boxplot

```{r}
data(iris)
my_colours <- c('firebrick', 'cornflowerblue', 'seagreen')

# Set graphical parameters:
# mar = c(bottom, left, top, right) margins
# bty  = "l" draws only the left and bottom plot border (bty = box type)
par(mar = c(4, 4, 2, 2), bty = "l")

# horizontal boxplots of sepal length for each species
boxplot(Sepal.Length ~ Species, 
        data = iris, 
        col = my_colours, 
        horizontal = TRUE)
```

### Density plots

```{r}
iris.split <- split(iris, iris$Species)
d.set <- density(iris.split$setosa$Sepal.Length) 
d.vers <- density(iris.split$versicolor$Sepal.Length) 
d.virg <- density(iris.split$virginica$Sepal.Length)
```

```{r}
par(mar = c(5, 4, 2, 10), bty = 'l')
plot(d.set, main = '', xlab = 'S. Length', xlim = range(iris$Sepal.Length))
lines(d.set, col = my_colours[1])
lines(d.vers, col = my_colours[2])
lines(d.virg, col = my_colours[3])

legend(8.2, 0.9, legend = unique(iris$Species), 
       title = 'SPECIES', 
       lwd = 1,
       xpd = TRUE, 
       col = my_colours)
```

# GGplot

1.  Data

```{r, fig.height=2, fig.width=2}
    ggplot(data = iris)
```

2.  Aesthetics

aesthetics: x and y position, grouping, colours, fills, shapes, area, etc.

```{{r, fig.height= 4, fig.width=4}}
ggplot(data = iris,
           mapping = aes(x = Species, y = Sepal.Length))
```
    

3.  geometrical layer

type of plot

```{r, fig.height = 4, fig.width = 4}
ggplot(data = iris,
       mapping = aes(x = Species, y = Sepal.Length)) +
  geom_boxplot()

```

4.  Extras

```{r, fig.height=5, fig.width=7}
    ggplot(data = iris,
           mapping = aes(x = Species, y = Sepal.Length)) +
      geom_boxplot(fill = my_colours) +
      coord_flip() +
      labs(y = "Sepal Length (mm)", 
           x = NULL, 
           title = "Something informative about this plot", 
           subtitle = "iris {datasets}") +
      theme_linedraw(base_size = 16)

```

### Density plot

```{r}
ggplot(data = iris,#fill here says to use Species to group and fill in by Species 
       mapping = aes(x = Sepal.Length, fill = Species)) +
  geom_density() +
  scale_fill_manual(values = my_colours)
  
```

```{r}
ggplot(data = iris,#fill here says to use Species to group and fill in by Species 
       mapping = aes(x = Sepal.Length, fill = Species)) +
  geom_density() +
  scale_fill_manual(values = my_colours) +
  facet_wrap(~Species)
```

## Exercises

Use `mtcars` to show the relationship between car weight (`wt`) and fuel economy (`mpg`). Make the plot deliberately ugly while keeping it technically valid. 

::: {callout-tip collapse="true"}

### Example solution
    
```{r fig.height= 6}
ggplot(mtcars, aes(x = wt, y = mpg,colour = factor(cyl), 
                   shape = factor(gear), size = hp)) +
  geom_point(alpha = 0.2) +
  scale_colour_manual(values = c("magenta", "yellow", "cyan")) +
  scale_size_continuous(range = c(3, 20)) +
  labs(title = "AN EXTREMELY IMPORTANT CAR GRAPH",
       x = "Weight",
       y = "MILES PER GALLON",
       colour = "CYLINDERS",
       shape = "GEARS",
       size = "HORSEPOWER")
```
    
:::


::: callout-tip
## Resources

-   [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) Claus O. Wilke
-   
:::
