# 0HM120 Advanced data analysis

## Week 1: Fundamentals of descriptive and inferential statistics

### Describing the population

```{r}
heights <- c(160, 170, 170, 170, 180, 180, 190, 200)
median(heights)
mean(heights)
```

`mode` in R returns the objects type. To get the statistical mode with base R you need to write a function for mode

```{r}
stat_mode <- function(x){
  tab <- table(x)
  mode_cha <- names(tab)[tab == max(tab)]# annoying it returns a character 
  mode <- as.numeric(mode_cha)
  return(mode)
}
stat_mode(heights)
```

```{r}
range(heights)
max(heights) - min(heights)
```

```{r}
var(heights)
sqrt(var(heights))
sd(heights)
```

range(), var(), and sd() summarize the data with individual numbers. A distribution shows the whole pattern of the data. For height, most people are near the average, while very short and very tall people are less common. So a probability distribution describes which values a variable (height) can be and how common or likely those values are.

```{r}
hist(heights, probability = TRUE,
     main = "Distribution of Heights",
     xlab = "Height")

curve(dnorm(x, mean = mean(heights), sd = sd(heights)),
      add = TRUE,
      col = "red",
      lwd = 2)
```

Taller bars represent more common ranges of height. The red curve is an idealised normal distribution with the same mean and standard deviation as our data. *Note* density: estimated probability per unit on the x-axis. For more on probability density functions (PDF) and kernel density estimation see: [The chemical statistician](https://chemicalstatistician.wordpress.com/2013/06/09/exploratory-data-analysis-kernel-density-estimation-in-r-on-ozone-pollution-data-in-new-york-and-ozonopolis/)

::: callout-tip

You can already see from the summary statistics averages (mean, median, and mode) and more clearly from the plot that sample of heights are not normally distributed. In a truly normal distribution the mean = median = mode. In practice they are close together. Here there are only 8 observations: with a small sample, there is a greater chance of getting an unrepresentative picture of the population because of random variation.

:::

```{ojs}
//| echo: false

mu = 180
sigma = 11

// Create one fixed simulated population
randomHeight = d3.randomNormal(mu, sigma)
population = d3
  .range(1000)
  .map(() => Math.round(randomHeight())) //nearest cm 

// Control: set your sample size 
viewof sampleSize = Inputs.range(
  [2, 100],
  {
    value: 8,
    step: 1,
    label: "Sample size"
  }
)

viewof newSample = Inputs.button("Draw another sample")

// Draw a new sample when either input changes
sample = {
  sampleSize;
  newSample;

  return d3
    .shuffle(population.slice())
    .slice(0, sampleSize);
}

// Create 10 cm histogram bins
bins = d3
  .bin()
  .domain([100, 200])
  .thresholds(d3.range(110, 200, 10))(sample)

// Convert bin counts to probability densities
histogram = bins.map(bin => ({
  x0: bin.x0,
  x1: bin.x1,
  density: bin.length /
    (sample.length * (bin.x1 - bin.x0))
}))

// Theoretical normal distribution
normalCurve = d3.range(100, 200.1, 0.25).map(x => ({
  x: x,
  density:
    Math.exp(-0.5 * ((x - mu) / sigma) ** 2) /
    (sigma * Math.sqrt(2 * Math.PI))
}))

// Current sample statistics
sampleMean = d3.mean(sample)
sampleSD = d3.deviation(sample)

// Interactive plot
Plot.plot({
  width: width,
  height: 430,

  title: `Sample size: ${sampleSize}`,

  subtitle:
    `Sample mean = ${sampleMean.toFixed(1)} cm; ` +
    `sample SD = ${sampleSD.toFixed(1)} cm`,

  caption:
    "Red curve and dashed line: population. Blue line: sample mean.",

  x: {
    domain: [100, 200],
    label: "Height (cm)"
  },

  y: {
    domain: [0, 0.1], // max possible is 1/binwidth
    label: "Density",
    grid: true
  },

  marks: [
    Plot.rectY(histogram, {
      x1: "x0",
      x2: "x1",
      y1: 0,
      y2: "density",
      fill: "lightgray",
      stroke: "white",
      inset: 1
    }),

    Plot.line(normalCurve, {
      x: "x",
      y: "density",
      stroke: "red",
      strokeWidth: 3
    }),

    // True population mean
    Plot.ruleX([mu], {
      stroke: "red",
      strokeWidth: 2,
      strokeDasharray: "5,4"
    }),

    // Current sample mean
    Plot.ruleX([sampleMean], {
      stroke: "blue",
      strokeWidth: 3
    }),

    Plot.ruleY([0])
  ]
})
```




